\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-09-23 chapters/bci.tex
Started paradigms section

\todo{
Locked-in -\> communicate
Paraplegic -\> walk again
tetraplegic, locked-in syndrome
Regaining capabilities and improving quality of life
Start from wolpaw examples, for spelling
Cite some catchy state of the art examples (walk again, speech avatar Metzger, handwriting
willet, neuralink)
}
\todo{put focus on communication BCI}

\section{A direct interface to the brain}
\comment{This section describes a bci in layman's terms, without too much
scientific background, jargon and literature, in the following sections of this
chapter a more scientific style is used}
A \ac{bci} reads your mind.
\ac{bci}s are an entertaining subject with many futuristic and more established
applications.
Devices processing direct inputs from the central nervous system can be
useful for rehabilitation, medical diagnosis or treatment.
They have always been especially promising as assistive technology for
paralyzed patients limited in their communication ability. \todo{expand more on this last
sentence, put patients first in introduction, see above}
Furthermore, they bring a new paradigm to human-computer interaction,
especially when paired with virtual or augmented reality.

@@

\section{Recording technologies}
\label{sec:bci-recording}

@@

Specifically, invasive methods come with an increased \emph{spatial resolution},
meaning they can extract a signal from a specific brain region or even a set of
neurons of interest.
Generally, penetrating microelectrodes are considered to have the highest
spatial resolution.
Microelectrodes can measure the Local Field Potential (LFP), the extracellular
potential of a small group of neurons, or even intracellular single neuron
action potential spikes.
They come in single-electrode form, or, more popularily, in microelectrode
arrays.
Well-known examples are the Utah array~\cite{Maynard1997} and the more recent
NeuraLink implant~\cite{Musk2019}.

@@

\Ac{eeg} is the de facto non-invasive electrophysiology standard, in clinical
neurology practice as well as in \ac{bci} research.
Developed in 1924, it is cheap, relatively practical and can offers great
temporal resolution up to thousands of Hz.
As it is easily applicable on the outside of the scalp, many practical
consumer grade \ac{eeg} headsets exist in the form (often wireless) electrode
caps or helmets.
Current consumer systems can feature dry electrodes, but clinical and
research systems often use electrolyte gel to reduce impedance for improved signal
quality, which slightly decreases practicality.
Nevertheless, it is the most accessible \ac{bci} technology for both patients
and researchers.

@@

\section{\ac{bci} Paradigms}
\label{sec:bci-paradigms}
\fullpagefig{figures/bci/paradigms.eps}{A \ac{bci} paradigm defines which brain
activity will be used for control. They can be categorized in \emph{active},
\emph{reactive} and \emph{passive} paradigms, depending on their reliance on
external stimulation and active participation required from the user.
Adapted from~\textcite{Muehl2014}.
}{fig:bci-paradigms}
\todo{check premissions}
\todo{cite A comprehensive review of ac{eeg}-based brain–computer interface
paradigms}
passive:
also monitoring for diagnostic purposes
emotion recognition
neurofeedback
error potentials
cannot easily be used for fast communication
synchronous <-> asynchronous

\cite{Xu2021}

@@

somatosensory evoked potentials (cite jimmy)

@@

Usually, a decoder makes a prediction for a given \emph{trial} while operating
the \ac{bci}.
A trial is the smallest unit on which a selection decision can be made, for
instance be one imagined movement, or one repetition of flashing all targets
in a visual \ac{erp} \ac{bci}.
Some valuable metrics to asses decoder performance in the classification case
are accuracy and \ac{rocauc}.
Accuracy can be obtained from the correctness of predicted trial selection as
\begin{equation}
  \text{accuracy} = \frac{\text{TP}+\text{TN}}{\text{TP}+\text{TN}+\text{FP}+\text{FN}}
\end{equation}
with TP and TN the numbers of true positives and negatives and FP and FP the
numbers of false positives and negatives.
\Ac{rocauc} is also a measure of classifier predictive power, but is more
suited for evaluation of classification of single epochs of data and in the
presence of imbalanced data, e.g. when
comparing single \ac{erps}.
Higer \ac{rocauc} usually translates to a higher target selection accuracy.

@@

\section{A case study: the visual oddball ERP paradigm}
good choice because it scores high on the parameters above
\subsection{Stimulation}
\todo{figure of P3 matrix}
\todo{cite P300 literature, first example and current sota}
\subsection{The event-related potential}
P300 and other components
does not only rely on P300, but also heavily dependent on visual early
components, hence we'll refer to it under the more generall oddball ERP
paradigm
\todo{figure of ERP components}
\subsection{Preprocessing and feature extraction}
referencing, filtering, epoching, baselining
\subsection{Decoding ERPs}
tLDA, beamforming
Linear decoding techniques

>>>>

\todo{
Locked-in -\> communicate
Paraplegic -\> walk again
tetraplegic, locked-in syndrome
Regaining capabilities and improving quality of life
Start from wolpaw examples, for spelling
Cite some catchy state of the art examples (walk again, speech avatar Metzger, handwriting
willet, neuralink)
}
\todo{put focus on communication BCI}
\section{A direct interface to the brain}
\comment{This section describes a bci in layman's terms, without too much
scientific background, jargon and literature, in the following sections of this
chapter a more scientific style is used}
A \ac{bci} reads your mind.
\ac{bci}s are an entertaining subject with many futuristic and more established
applications.
Devices processing direct inputs from the central nervous system can be
useful for rehabilitation, medical diagnosis or treatment.
They have always been especially promising as assistive technology for
paralyzed patients limited in their communication ability. \todo{expand more on this last
sentence, put patients first in introduction, see above}
Furthermore, they bring a new paradigm to human-computer interaction,
especially when paired with virtual or augmented reality.

@@

\section{Recording technologies}
\label{sec:bci-recording}
\todo{make sure every part ends with a motivation for why we chose our
approach}

@@

Specifically, invasive methods come with an increased \emph{spatial resolution},
meaning they can extract a signal from a specific brain region or even a set of
neurons of interest.
Generally, penetrating microelectrodes are considered to have the highest
spatial resolution.
Microelectrodes can measure the Local Field Potential (LFP), the extracellular
potential of a small group of neurons, or even intracellular single neuron
action potential spikes.
They come in single-electrode form, or, more popularly, in microelectrode
arrays.
Well-known examples are the Utah array~\cite{Maynard1997} and the more recent
NeuraLink implant~\cite{Musk2019}.

@@

\Ac{eeg} is the de facto non-invasive electrophysiology standard, in clinical
neurology practice as well as in \ac{bci} research.
Developed in 1924~\cite{Berger1929}, it is cheap and relatively practical and
can offer great temporal resolution up to thousands of Hz.
As it is easily applicable on the outside of the scalp, many practical
consumer grade \ac{eeg} headsets exist in the form (often wireless) electrode
caps or helmets.
Current consumer systems can feature dry electrodes, but clinical and
research systems often use electrolyte gel to reduce impedance for improved signal
quality, which slightly decreases practicality.
Nevertheless, it is the most accessible \ac{bci} technology for both patients
and researchers.

@@

\section{\ac{bci} Paradigms}
\label{sec:bci-paradigms}

The \ac{bci} paradigm~\cite{Xu2021} is the key to translating brain activity into useful
output commands, since it defines which neural responses will be elicited and
captured.
A paradigm can be loosely seen as \emph{a way of interacting with the
\ac{bci}}.
According to \textcite{Zander2011}, they can be arranged by their reliance on
external stimulation and the engagement of the user as shown in
Figure~\ref{fig:bci-paradigms}.

\fullpagefig{figures/bci/paradigms.eps}{
  A \ac{bci} paradigm defines which brain activity will be used for control.
  They can be categorized in \emph{active}, \emph{reactive} and \emph{passive}
  paradigms, depending on their reliance on external stimulation and active
  participation required from the user.
  Adapted from~\textcite{Muehl2014}.
}{fig:bci-paradigms}

Paradigms that require little active participation of the user like
redirecting attention or initiating imagined actions are classified as passive
\ac{bci}.
From the user's point of view, this concept is probably the most attractive,
since their intentions or cognitive state state would be inferred from just
their brain activity.
However, without active user participation, it can be hard to establish reliable
communication or control.
Therefore, passive \ac{bci} are currently more suited to tasks like
emotion or affect monitoring~\cite{Torres2020,Libert2019}, workload
monitoring~\cite{Zanetti2021} or, more dependent on stimulation, error
detection~\cite{SiMohammed2020}.

Furthermore, active paradigms make it easier to quickly collect training data for
supervized machine learning since the conditions can be more easily controlled
by the user or the stimulation.

active

reactive
works for most people
high effort, unintuitive and low expressive power compared to movement and
speech~\cite{Zander2011}, which are more natural and can be emulated with active bci

\todo{check permissions}
\todo{cite A comprehensive review of ac{eeg}-based brain–computer interface
paradigms}
passive:
also monitoring for diagnostic purposes
emotion recognition
neurofeedback
error potentials
cannot easily be used for fast communication
synchronous <-> asynchronous
Again, asynchronous is preferred by users because it is more intuitive and
comfortable to use, but it is far easier to decode synchronous responses since
it is known when to expect a response and when a decision should be made.

@@

somatosensory evoked potentials~\cite{Petit2021} (cite jimmy)

@@

Usually, a decoder makes a prediction for a given \emph{trial} while operating
the \ac{bci}.
A trial is the smallest unit on which a selection decision can be made, for
instance be one imagined movement, or one repetition of flashing all targets
in a visual \ac{erp} \ac{bci}.
Some valuable metrics to asses decoder performance in the classification case
are accuracy and \ac{rocauc}.
Accuracy can be obtained from the correctness of predicted trial selection as
\begin{equation}
  \text{accuracy} = \frac{\text{TP}+\text{TN}}{\text{TP}+\text{TN}+\text{FP}+\text{FN}}
\end{equation}
with TP and TN the numbers of true positives and negatives and FP and FP the
numbers of false positives and negatives.
\Ac{rocauc} is also a measure of classifier predictive power, but is more
suited for evaluation of classification of single epochs of data and in the
presence of imbalanced data, e.g. when
comparing single \acp{erp}.
Higer \ac{rocauc} usually translates to a higher target selection accuracy.

@@

\section{A case study: the visual oddball ERP paradigm}
good choice because it scores high on the parameters above
\subsection{Stimulation}
comfortable, preferred over SSVEP, similar ITR is this true?
\todo{figure of P3 matrix}
\todo{cite P300 literature, first example and current sota}
\subsection{The event-related potential}
P300 and other components
does not only rely on P300, but also heavily dependent on visual early
components, hence we'll refer to it under the more generall oddball ERP
paradigm
\todo{figure of ERP components}
\subsection{Preprocessing and feature extraction}
referencing, filtering, epoching, baselining
\subsection{Decoding ERPs}
tLDA, beamforming
Linear decoding techniques

### 2024-09-23 chapters/cv.tex
Recording technologies
Started paradigms section

\subsection*{About me}
I am a Ph.D. candidate\todo{change} at KU Leuven Laboratory for Neuro- and Psychophysiology
and the University of Lille Research Center for Informatics, Signal Processing,
and Control Science. My research is centered on developing brain-computer
interface (BCI) technologies with a primary focus on assisting individuals with
severe disabilities, especially those with motor and eye movement impairments.
My goal is to enhance their quality of life by creating accurate and
user-friendly BCIs. Combining my expertise in machine learning, signal
processing, and user interface design with insights from neurophysiology, my
PhD centers around the development of an event-related potential BCI system
using EEG for real-time communication. I strongly believe in the potential of
innovative neurotechnology to empower individuals and patients, enabling them
to lead more independent and fulfilling lives.

@@

\subsubsection{Master theses and internships supervised}
\begin{cvlist}
	\cvitem{Eye-tracker and ERP data fusion for gaze-independent visual
		BCI}{\ \\ Juliette  Meunier, University of Lille, 2024}
	\cvitem{LDA in	the combined
		space-time-frequency domain\\ for BCI decoding}{\ \\Lunkyadi Sucipto, KU Leuven,
		2023--2023}
	\cvitem{Tackling the Midas Touch problem in eye tracking with a
		BCI}{\ \\Reniflal Ebenezer Sundaralal, KU Leuven, 2023--2023}
	\cvitem{An EEG connectivity analysis of Alzheimer’s Disease
		and Frontotemporal Dementia}{Zoe Barinaga, KU Leuven, 2023--2024}
	\cvitem{Single-trial ERP latencies as a predictor for the mode
		of visual attention}{\ \\Yildiz Dilara Parry, KU Leuven, 2023}
	\cvitem{A hybrid P300-gaze BCI alternative for navigating virtual
		spaces}{\ \\Gijs Claes, KU Leuven, 2021--2022}
\end{cvlist}

>>>>

\subsection*{About me}
I am a Ph.D. candidate\todo{change} at KU Leuven Laboratory for Neuro- and Psychophysiology
and the University of Lille Research Center for Informatics, Signal Processing,
and Control Science. My research is centered on developing brain-computer
interface (BCI) technologies with a primary focus on assisting individuals with
severe disabilities, especially those with motor and eye movement impairments.
My goal is to enhance their quality of life by creating accurate and
user-friendly BCIs. Combining my expertise in machine learning, signal
processing, and user interface design with insights from neurophysiology, my
Ph.D. centers around the development of an event-related potential BCI system
using EEG for real-time communication. I strongly believe in the potential of
innovative neurotechnology to empower individuals and patients, enabling them
to lead more independent and fulfilling lives.

@@

\subsubsection{Master theses and internships supervised}
\begin{cvlist}
	\cvitem{Eye-tracker and ERP data fusion for gaze-independent visual
		BCI}{\ \\ Juliette  Meunier, University of Lille, 2024}
	\cvitem{LDA in	the combined
		space-time-frequency domain for BCI decoding}{\ \\Lunkyadi Sucipto, KU Leuven,
		2023--2023}
	\cvitem{Tackling the Midas Touch problem in eye tracking with a
		BCI}{\ \\Reniflal Ebenezer Sundaralal, KU Leuven, 2023--2023}
	\cvitem{An EEG connectivity analysis of Alzheimer’s Disease
		and Frontotemporal Dementia}{Zoe Barinaga, KU Leuven, 2023--2024}
	\cvitem{Single-trial ERP latencies as a predictor for the mode
		of visual attention}{\ \\Yildiz Dilara Parry, KU Leuven, 2023}
	\cvitem{A hybrid P300-gaze BCI alternative for navigating virtual
		spaces}{\ \\Gijs Claes, KU Leuven, 2021--2022}
\end{cvlist}

### 2024-09-23 chapters/gaze_independence.tex
Started paradigms section

One of the goals of Brain-computer interfacing is establishing a communication
channel that does not rely on speech or muscular activity~\cite{Naci2012,Chaudhary2016},
which in turn can provide solutions to paralyzed individuals.
In the strictest interpretation, this also means that an interface should not
rely on the control of eye muscles used to redirect the gaze or for blinking.
\todo{patients suffering paralysis due to damage to the CNS (stroke, spinal
chord injury, tbi) or neurodegenerative diseases}

In fact, many patients in the BCI target population can suffer to some extent
from eye motor disability, requiring BCIs adapted to their condition.
Table~\ref{tab:incidence} reports the relatively high frequencies of
eye motor problems, which can range from minor (nystagmus
\footnote{
Involuntary, rhythmic, and repetitive eye movements recognizable by their
consistent directionality (horizontal, vertical, or rotational).
}, other eye tremors
\footnote{
These can include square-wave jerks, saccadic intrusions, microtremors or
microsaccades while resting or fixating the gaze.
}, gaze fixation fatigue or discomfort, \ldots), to severe (partial ophtalmoplegia
\footnote{
Weakness or limited paralysis of one or more of the muscles that control eye
movement, leading to restricted eye motion, but not complete paralysis.
Often, a specific movement direction (up-down, left-right) is preserved.
}, involuntary movements, impaired pursuit, \ldots) and even complete
ophtalmoplegia\footnote{Full eye movement paralysis.} or eye motor paresis.
This not only has an effect on the patients' vision and coordination, but also
on their ability to operate a visual \ac{bci}.

\begin{table}[h]
  \sffamily
  \footnotesize
	\centering
	\begin{tabular}{@{}l|rrrrr|r@{}}
              & \bfseries ALS  & \bfseries MS   & \bfseries Stroke  &\bfseries DMD &\bfseries SMA  &\bfseries LiS  \\ \hline
		\bfseries Minor    & 50\% & 31\% & 40-70\% & + & - &      \\
		\bfseries Severe   & 33\% & 3\%  & +       & - & - & 98\% \\
		\bfseries Complete & 17\% & -    & +       & - & - & 2\%  \\
	\end{tabular}
	\caption{Incidence of eye motor impairment in selected BCI target
		patient populations. (ALS: Amyotrophic Lateral Sclerosis, MS: Multiple Sclerosis,
		DMD: Duchenne's Muscular Dystrophy, SMA: Spinal Muscular Atrophy, LiS:
		Locked-In-Syndrome). $+$: frequent, $-$: infrequent.}\label{tab:incidence}
\end{table}
\todo{include FA}

Among the most affected patients are stroke patients~\cite{Pollock2011}, mostly
due to brain-stem or cerebellar stroke~\cite{Moncayo2009,Bogousslavsky1987}.
\todo{summarize and cite Rowe2019}
These patients can suffer from severe or complete eye motor impairment immediately
immediately from the onset of their condition.
In \ac{als}, a progressive disease, oculomotor issues are also fairly common.
Although eye movement is often cited as one of the longest preserved
capabilities in \ac{als}, \textcite{Guo2022} show that minor issues are still
fairly common,
especially for patients with bulbar onset \ac{als}.
Furthermore \textcite{Hayashi1991} show that, as these patients are supported to beyond
respiratory failure, the disease progression will eventually also involve
paralysis of the eye muscles.
Various forms of eye movement abnormalities also occur often and are especially well studied in
\ac{ms}~\cite{Mueri1985,Prasad2010,Castelnovo2016,Serra2018,Polet2020}.
These abnormalities can be minor or severe, seldom progressing to complete
paralysis.
However, \ac{ms} often comes with vision loss, further complicating interaction
with visual \acp{bci}
\todo{FA}
Patients from other neurodegenerative diseases like
\ac{sma}~\cite{Anagnostou2021} and \ac{dmd}~\cite{Lui2001} are
sometimes also interested in \ac{bci} use, but their eye motor capabilities
mostly preserved.

\newcommand\fnlis{\footnote{
Multiple definitions of \ac{lis} are encountered in
BCI and neurological literature.
Some definitions include only tetraplegic patients without eye movements used
for communications.
Others distinguish Complete Locked-in Syndrome (CLIS) with full body paralysis,
including no eye motor control at all, from a \ac{lis} state with some preserved eye
movements or minor motor output.
While some definitions only include patients who suffered stroke or traumatic
brain injury to specific regions in the brain (midbrain, brainstem,
cerebellum)~\cite{Smith2005},
it can also generally refer to the state of full body paralysis
or loss of muscle tone incurred in neurodegenerative diseases, combined with
the inability to speak, such as occurs in late stage ALS.
In clinical reality, every tetraplegic patient has their own unique set of
preserved capabilities and from an solution-oriented \ac{bci} engineering point
of view, the etiology of the symptoms can be abstracted away.
}}

\newcommand\fnwolpawcrit{\footnote{
\it``The first class consists of people who are truly totally locked-in (e.g.,
due to end-stage ALS or severe cerebral palsy), who have no remaining
useful neuromuscular control of any sort, including no eye movement.
\elide\ This class is very small. \elide\
The second class of potential BCI users comprises those who retain
a very limited capacity for neuromuscular control. This group includes
people who retain some useful eye movement or enough limb muscle
function to operate a single-switch system. Such control is often slow,
unreliable, or easily fatigued.
This group is much larger than the first.
\elide\
The third class of potential BCI users, which is the largest of all,
includes those who still retain (and can be expected to continue to
retain) substantial neuromuscular control, particularly speech and/or
hand control, and can, therefore, operate a wide range of assistive com-
munication and control devices.''
}}

Together, some of these patients constitute the \ac{lis} patient group, which
forms one of the main \ac{bci} target populations.
In this work, we'll define \ac{lis} patients as those a situation of (near)
complete paralysis and difficulties or the inability to communicate without
assistive technology, corresponding corresponds to classes one and two defined by
\textcite{Wolpaw2006}\fnwolpawcrit.
Additionally some degree of severe or complete eye motor impairment is ususally
necessary to qualify as locked-in\fnlis.
Since this work is concerned with gaze-impaired patients, we will adopt this
interpretation.
Case study reports~\cite{Patterson1986} show that the population of locked-in
patients with complete eye motor paralysis is very small.
Hence, it would be interesting to focus efforts in \ac{bci} development on the
larger population of \ac{lis} patients that  currently slip through the cracks
of the current assistive technology offer.
These are the patients that require assistance to communicate independently,
which is crucial to retaining an acceptable quality of life, but whose severe
eye motor impairment prevents them from using eye tracking based solutions.

\todo{order of concepts in next two paragraphs can be tweaked}
This leads us to what we'll call the \emph{gaze-dependence problem in visual
BCI}:
Traditional visual BCI scenarios require the user to overtly direct their both
\ac{vsa} and gaze toward the screen target they intend to select.
For our purposes, screen targets are overlaid with non-overlapping, transient
stimuli that evoke event-related  potentials \acp{erp} in the \ac{eeg}.
The selected target can be decoded from these \acp{erp}, as is the case for the
oddball paradigm where observing a rare but attended stimulus evokes a P3 ERP
component.
However, a critical challenge arises when users rely solely or in part on covert
VSA, which involves directing visuospatial attention without corresponding eye gaze.
In these cases, classical solutions often fall short of the widely accepted
80\% target selection accuracy threshold~\cite{Brunner2010,Frenzel2011,Treder2010,Ron2019} deemed necessary for a comfortable user
experience~\cite{Neeling2019}, calling for alternative, gaze-independent
solutions.

@@

This leads to an often unadressed paradox: the BCI target population consists
of patients who suffer from pathologies like Amyotrophic Lateral Sclerosis
(ALS), Multiple Sclerosis (MS), stroke (specifically
brain-stem stroke) or traumatic brain injury, which can lead to several degrees
of paralysis, or even to Locked-in Syndrome (LiS), the complete loss of muscle
control with preserved consciousness.
While BCIs are most attractive as a solution for these patients
compared to other assistive technologies like eye-tracking, studies often
report bad performance in patient populations precisely due to the lack of adequate eye motor
control, or patients with gaze impairments are excluded.
Different degrees of eye motor impairment can render visual BCIs
uncomfortable or outright impossible to use.
When operating a visual BCI, the usual rapid series of forced saccades followed
by fixation is tiring over time, even for healthy control subjects.
A suitable alternative would allow the user to keep their eyes in an at-rest
position of their choice while operating the BCI.
This points to the need for gaze-independent BCIs.

>>>>

\todo{Algemeen: in begin uitleggen dat de ziektes er niet veel toe doen en dat
het gaat over verlamde personen, verlies van spierkracht en coördinatie en dus
gehinded in communicatie, helemaal in begin van chapter 1 duidelijk maken dat
het om deze patiënten gaat}
One of the goals of brain-computer interfacing is establishing a communication
channel that does not rely on speech or muscular activity~\cite{Naci2012,Chaudhary2016},
which in turn can provide solutions to paralyzed individuals.
In the strictest interpretation, this also means that an interface should not
rely on the control of eye muscles used to redirect the gaze or for blinking.
\todo{patients suffering paralysis due to damage to the CNS (stroke, spinal
chord injury, tbi) or neurodegenerative diseases}

In fact, many patients in the BCI target population suffer to some extent
from eye motor disability, requiring BCIs adapted to their condition.
Table~\ref{tab:incidence} reports the relatively high frequencies of
eye motor problems, which can range from minor (nystagmus
\footnote{
Involuntary, rhythmic, and repetitive eye movements recognizable by their
consistent directionality (horizontal, vertical, or rotational).
}, other eye tremors
\footnote{
These can include square-wave jerks, saccadic intrusions, microtremors, or
microsaccades while resting or fixating the gaze.
}, gaze fixation fatigue or discomfort, \ldots), to severe (partial ophtalmoplegia
\footnote{
Weakness or limited paralysis of one or more of the muscles that control eye
movement, leading to restricted eye motion, but not complete paralysis.
Often, a specific movement direction (up-down, left-right) is preserved.
}, involuntary movements, impaired pursuit, \ldots), and even complete
ophtalmoplegia\footnote{Full eye movement paralysis.} or eye motor paresis.
This not only has an effect on vision and coordination but also
on their ability to operate a visual \ac{bci}.

\begin{table}[h]
  \sffamily
  \footnotesize
	\centering
	\begin{tabular}{@{}l|rrrrr|r@{}}
    & \bfseries \acs{als} & \bfseries \acs{ms}   & \bfseries Stroke &\bfseries\acs{dmd} &\bfseries \acs{sma} &\bfseries \acs{lis} \\ \hline
		\bfseries Minor    & 50\% & 31\% & 40-70\% & + & - &      \\
		\bfseries Severe   & 33\% & 3\%  & +       & - & - & 98\% \\
		\bfseries Complete & 17\% & -    & +       & - & - & 2\%  \\
	\end{tabular}
	\caption{Incidence of eye motor impairment in selected BCI target
    patient populations. (\acs{als}: \acl{als}, \acs{ms}: \acl{ms}, \acs{dmd}:
    \acl{dmd}, \acs{sma}: \acl{sma}, \acs{lis}: \acl{lis}).
    $+$: frequent, $-$: infrequent.}
    \label{tab:incidence}
\end{table}
\todo{include FA}

Among the most affected patients are stroke patients~\cite{Pollock2011}, mostly
due to brainstem or cerebellar stroke~\cite{Moncayo2009,Bogousslavsky1987}.
\todo{summarize and cite Rowe2019}
These patients might suffer from severe or complete eye motor impairment
immediately from the onset of their condition.
In \ac{als}, a progressive disease, oculomotor issues are also fairly common.
Although eye movement is often cited as one of the longest preserved
capabilities in \ac{als}, \textcite{Guo2022} show that minor issues are still
fairly common,
especially for patients with bulbar onset \ac{als}.
Furthermore, \textcite{Hayashi1991} show that, as these patients are supported beyond
respiratory failure, the disease progression will eventually also involve
paralysis of the eye muscles.
Various forms of eye movement abnormalities also occur often and are especially well studied in
\ac{ms}~\cite{Mueri1985,Prasad2010,Castelnovo2016,Serra2018,Polet2020}.
These abnormalities can be minor or severe, seldom progressing to complete
paralysis.
However, \ac{ms} often comes with vision loss, further complicating interaction
with visual \acp{bci}.
\todo{FA}
Patients suffering from other neurodegenerative diseases like
\ac{sma}~\cite{Anagnostou2021} and \ac{dmd}~\cite{Lui2001} are
sometimes also interested in \ac{bci} use, but their eye motor capabilities are
mostly preserved.

\newcommand\fnlis{\footnote{
Multiple definitions of \ac{lis} are encountered in
BCI and neurological literature.
Some definitions include only tetraplegic patients without eye movements used
for communications.
Others distinguish Complete Locked-in Syndrome (CLIS) with full body paralysis,
including no eye motor control at all, from a \ac{lis} state with some preserved eye
movements or minor motor output.
While some definitions only include patients who suffered stroke or traumatic
brain injury to specific regions in the brain (midbrain, brainstem, or
cerebellum)~\cite{Smith2005},
it can also generally refer to the state of full body paralysis
or loss of muscle tone incurred in neurodegenerative diseases, combined with
the inability to speak, such as occurs in late stage ALS.
In clinical reality, every tetraplegic patient has their own unique set of
preserved capabilities and from an solution-oriented \ac{bci} engineering point
of view, the etiology of the symptoms can be abstracted away.
}}

\newcommand\fnwolpawcrit{\footnote{
\it``The first class consists of people who are truly totally locked-in (e.g.,
due to end-stage ALS or severe cerebral palsy), who have no remaining
useful neuromuscular control of any sort, including no eye movement.
\elide\ This class is very small. \elide\
The second class of potential BCI users comprises those who retain
a very limited capacity for neuromuscular control. This group includes
people who retain some useful eye movement or enough limb muscle
function to operate a single-switch system. Such control is often slow,
unreliable, or easily fatigued.
This group is much larger than the first.
\elide\
The third class of potential BCI users, which is the largest of all,
includes those who still retain (and can be expected to continue to
retain) substantial neuromuscular control, particularly speech and/or
hand control, and can, therefore, operate a wide range of assistive
communication and control devices.''
}}

Together, some of these patients constitute the \ac{lis} patient group, which
forms one of the main \ac{bci} target populations.
In this work, we'll define \ac{lis} patients as those in a situation of (near)
complete paralysis and difficulties or the inability to communicate without
assistive technology
This corresponds to classes one and two defined by
\textcite{Wolpaw2006}\fnwolpawcrit.
Additionally, some degree of severe or complete eye motor impairment is usually
necessary to qualify as locked-in\fnlis.
Since this work is concerned with gaze-impaired patients, we will adopt this
interpretation into our definition.
Case study reports~\cite{Patterson1986} show that the population of locked-in
patients with complete eye motor paralysis is very small.
Hence, it would be interesting to focus \ac{bci} development efforts on the
larger population of \ac{lis} patients with severe eye motor impairment that
currently slip through the cracks of the assistive technology offer.
These are the patients whose severe eye motor impairment prevents them from
using eye tracking based solutions.
They currently use their remaining motor control to
communicate by indicating symbols on a letterboard with great effort,
or signal with upwards eye movements or blinks to confirm prompted letters.
They require a caregiver or relative to interpret their signals, and crave
the ability to communicate independently which is crucial to retaining an
acceptable quality of life.
\todo{order of concepts in next two paragraphs can be tweaked}

This leads us to what we'll call the \emph{gaze-dependence problem in visual
BCI}:
Traditional visual BCI scenarios require the user to overtly direct their both
\ac{vsa} and gaze toward the screen target they intend to select.
For our purposes, screen targets are overlaid with non-overlapping, transient
stimuli that evoke event-related  potentials \acp{erp} in the \ac{eeg}.
The selected target can be decoded from these \acp{erp}, as is the case for the
oddball paradigm where observing a rare but attended stimulus evokes a P3 ERP
component.
However, a critical challenge arises when users rely solely or in part on covert
VSA, which involves directing visuospatial attention without corresponding eye gaze.
In these cases, classical solutions often fall short of the widely accepted
80\% target selection accuracy threshold~\cite{Brunner2010,Frenzel2011,Treder2010,Ron2019} deemed necessary for a comfortable user
experience~\cite{Neeling2019}, calling for alternative, gaze-independent
solutions.

@@

This leads to the often unadressed gaze-dependence paradox: the BCI target
population consists of patients who suffer from various degrees
of paralysis, or even from Locked-in Syndrome (LiS), the complete loss of muscle
control with preserved consciousness.
While BCIs are most attractive as a solution for these patients
compared to other assistive technologies like eye-tracking, studies often
report bad performance in patient populations precisely due to the lack of adequate eye motor
control, or patients with gaze impairments are excluded.
Different degrees of eye motor impairment can render visual BCIs
uncomfortable or outright impossible to use.
When operating a visual BCI, the usual rapid series of forced saccades followed
by fixation is tiring over time, even for healthy control subjects.
A suitable alternative would allow the user to keep their eyes in an at-rest
position of their choice while operating the BCI.
This points to the need for gaze-independent BCIs.
